Modeling the effect of aneuploidy on cancer evolution

### 2023-02-06 EvolutionaryRescue1.tex
Add files via upload

@@

The probability of evolutionary rescue is given by:
\begin{equation}
p_{est}\sim\left\{
  \begin{array}{@{}ll@{}}
  1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right],\quad\text{if }4\lambda_avp_m>\left(\Delta_a-v\right)^2,\\
   1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right],\quad\text{if }\Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2,\\
   1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right],\quad\text{if }\Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2.
  \end{array}\right.
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{Logistic}
%We want to have birth and death rates which depends on the population size of wildtype $w$, aneuploidy $a$ and mutant $m$ cells:
%\begin{align*}
%&\lambda_w'=\lambda_w,\quad\mu_w'=\mu_w\\
%&\lambda_a'=C_1+\lambda_a\left(1-\frac{w+a+m}{K}\right),\quad \mu_a'=C_1\\
%&\lambda_m'=C_2+\lambda_m\left(1-\frac{w+a+m}{K}\right),\quad \mu_m'=C_2
%\end{align*}
%where $C_1, C_2>0$ are constants.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Standing genetic variation}
Until now we have assumed that the initial population of cells consisted entirely of wildtype cells. We modify this assumtion such that the initial population includes a fraction $f$ of cells with aneuploidy. The probability of evolutionary rescue by the cells with aneuploidy from the initial population is:
\begin{equation*}
p_{old}=1-\left(1-p_a\right)^{fN}\approx 1-\e^{-fNp_a}.
\end{equation*}
The total probability of evolutionary rescue is given by:
\begin{align}\nonumber
p_{total}&=p_{new}+\left(1-p_{new}\right)p_{old}\\
&=1-\e^{-\left[\left(1-f\right)p_w+fp_a\right]N}.
\end{align}
The fraction of the cases in which the population is rescued by the standing genetic variation is given by:
\begin{align*}
F\left(f\right)=\frac{p_{old}}{p_{total}}=\frac{ 1-\e^{-fNp_a}}{1-\e^{-\left[\left(1-f\right)p_w+fp_a\right]N}}.
\end{align*}
We let $F=\frac{1}{2}$ and we obtain:
\begin{equation}\label{halfeqstandvar}
f^*\approx\frac{p_w}{p_w+p_a},
\end{equation}
wher we have used the expansion $\e^x\approx 1+x$.
We plot $F$ and $f^*$ in Figure \ref{FractionPlot}.

@@

\begin{align}
\tau=\tau_1+\tau_2
\end{align}
inhomogeneus Poisson process
\begin{align}
\tau_1\sim \text{Poisson}\left(\frac{1}{up_aW\left(t\right)}\right)
\end{align}
where 
\begin{equation}
W\left(t\right)=W_0\e^{\Delta_wt}
\end{equation}
\begin{align}
f\left(\tau_1\right)=R\left(\tau_1\right)\e^{-\int_0^{\tau_1} R(t)\d t}
\end{align}
\begin{align*}
P\left(\tau_1<t\right)&=P\left(\tau_1<t|M\left(t\rightarrow\infty\right)\neq0\right)P\left(M\left(t\rightarrow\infty\right)\neq0\right)\\
&+P\left(\tau_1<t|M\left(t\rightarrow\infty\right)=0\right)P\left(M\left(t\rightarrow\infty\right)=0\right)\\
P\left(\tau_1<t\right)&=P\left(\tau_1<t|M\left(t\rightarrow\infty\right)\neq0\right)P\left(M\left(t\rightarrow\infty\right)\neq0\right)
\end{align*}
\begin{align}
P\left(\tau_1<t|M\left(t\rightarrow\infty\right)\neq0\right)=\frac{P\left(\tau_1<t\right)}{1-\left(1-p_w\right)^N}
\end{align}
where we have used
\begin{equation}
P\left(M\left(t\rightarrow\infty\right)\neq0\right)=1-\left(1-p_w\right)^N
\end{equation}
\begin{align}
f\left(\tau_1|M\left(t\rightarrow\infty\right)\neq0\right)=\frac{R\left(\tau_1\right)\e^{-\int_0^{\tau_1} R(t)\d t}}{1-\left(1-p_w\right)^N}
\end{align}
\begin{align}
\mathbb{E}\left[\tau_1\right]&=\frac{\int_0^\infty\e^{-\int_0^\tau R(t)\d t}\d\tau}{1-\left(1-p_w\right)^N}\\
&=\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}{1-\left(1-p_w\right)^N}
\end{align}
\begin{align}
P_2\left(\tau_2\right)&=\int_0^\infty \d w P\left(W_l=w\right)\left[1-\e^{-vp_mW_l\left(\tau_2\right)}\right]\\
&=1-\mathbb{E}\left[\e^{-vp_mW_l\left(\tau_2\right)}\right]
\end{align}
\begin{align}
a_\pm=\frac{2-\delta-y\pm\sqrt{\left(\delta+y\right)^2-4y}}{2\left(1-\delta\right)}
\end{align}
\begin{align}
P_2\left(\tau_2\right)=\frac{\left(a_+-1\right)\left(1-a_-\right)\left(1-\e^{\left[-\left(1-\delta\right)\left(a_+-a_-\right)t\right]}\right)}{a_+-1+\left(1-a_-\right)\e^{\left[-\left(1-\delta\right)\left(a_+-a_-\right)t\right]}}
\end{align}
with $y=vp_m$ in $a_\pm$.
\begin{align}
\mathbb{E}\left[\tau_2\right]=\int_0^\infty\left(1-\frac{P_2\left(t\right)}{p_m}\right)\d t
\end{align}
\section{Discussion}

>>>>

Most studies on evolutionary rescue are interested in the probability that at least one mutation rescues the population but the question of wheter multiple mutations contribute to the survival of the population is less explored. One exception is \cite{wilson2017soft} which show that evolutionary rescue is significantly inhanced by soft selective sweeps when multiple mutations contribute.

Evolutionary rescue which requires two succesive mutations has been investigated by \cite{martin2013probability} with the help of diffusion approximation.

@@

The probability of evolutionary rescue is given by:
\begin{equation}
p_{est}\sim\left\{
  \begin{array}{@{}ll@{}}
  1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right],\quad\text{if }4\lambda_avp_m>\left(\Delta_a-v\right)^2,\\
   1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right],\quad\text{if }\Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2,\\
   1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right],\quad\text{if }\Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2.
  \end{array}\right.
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Logistic}
We want to have birth and death rates which depends on the population size of wildtype $w$, aneuploidy $a$ and mutant $m$ cells:
\begin{align*}
&\lambda_w'=\lambda_w,\quad\mu_w'=\mu_w\\
&\lambda_a'=C_1+\lambda_a\left(1-\frac{w+a+m}{K}\right),\quad \mu_a'=C_1\\
&\lambda_m'=C_2+\lambda_m\left(1-\frac{w+a+m}{K}\right),\quad \mu_m'=C_2
\end{align*}
where $C_1, C_2>0$ are constants.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Standing genetic variation}
Until now we have assumed that the initial population of cells consisted entirely of wildtype cells. We modify this assumtion such that the initial population includes a fraction $f$ of cells with aneuploidy. The probability of evolutionary rescue by the cells with aneuploidy from the initial population is:
\begin{equation*}
p_{old}=1-\left(1-p_a\right)^{fN}\approx 1-\e^{-fNp_a}.
\end{equation*}
The total probability of evolutionary rescue is given by:
\begin{align}\nonumber
p_{total}&=p_{new}+\left(1-p_{new}\right)p_{old}\\
&=1-\e^{-\left[\left(1-f\right)p_w+fp_a\right]N}.
\end{align}
The fraction of the cases in which the population is rescued by the standing genetic variation is given by:
\begin{align*}
F\left(f\right)=\frac{p_{old}}{p_{total}}=\frac{ 1-\e^{-fNp_a}}{1-\e^{-\left[\left(1-f\right)p_w+fp_a\right]N}}.
\end{align*}
We let $F=\frac{1}{2}$ and we obtain:
\begin{equation}\label{halfeqstandvar}
f^*\approx\frac{p_w}{p_w+p_a},
\end{equation}
wher we have used the expansion $\e^x\approx 1+x$.
We plot $F$ and $f^*$ in Figure \ref{FractionPlot}.

@@

\begin{align}
\tau=\tau_1+\tau_2
\end{align}
inhomogeneus Poisson process
\begin{align}
\tau_1\sim \text{Poisson}\left(\frac{1}{up_aW\left(t\right)}\right)
\end{align}
where 
\begin{equation}
W\left(t\right)=W_0\e^{\Delta_wt}
\end{equation}
\begin{align}
f\left(\tau_1\right)=R\left(\tau_1\right)\e^{-\int_0^{\tau_1} R(t)\d t}
\end{align}
\begin{align*}
P\left(\tau_1<t\right)&=P\left(\tau_1<t|M\left(t\rightarrow\infty\right)\neq0\right)P\left(M\left(t\rightarrow\infty\right)\neq0\right)\\
&+P\left(\tau_1<t|M\left(t\rightarrow\infty\right)=0\right)P\left(M\left(t\rightarrow\infty\right)=0\right)\\
P\left(\tau_1<t\right)&=P\left(\tau_1<t|M\left(t\rightarrow\infty\right)\neq0\right)P\left(M\left(t\rightarrow\infty\right)\neq0\right)
\end{align*}
where we used
\begin{align*}
P\left(\tau_1<t|M\left(t\rightarrow\infty\right)=0\right)=0
\end{align*}
\begin{align}
P\left(\tau_1<t|M\left(t\rightarrow\infty\right)\neq0\right)=\frac{P\left(\tau_1<t\right)}{1-\left(1-p_w\right)^N}
\end{align}
where we have used
\begin{equation}
P\left(M\left(t\rightarrow\infty\right)\neq0\right)=1-\left(1-p_w\right)^N
\end{equation}
\begin{align}
f\left(\tau_1|M\left(t\rightarrow\infty\right)\neq0\right)=\frac{R\left(\tau_1\right)\e^{-\int_0^{\tau_1} R(t)\d t}}{1-\left(1-p_w\right)^N}
\end{align}
\begin{align}
\mathbb{E}\left[\tau_1\right]&=\frac{\int_0^\infty\e^{-\int_0^\tau R(t)\d t}\d\tau}{1-\left(1-p_w\right)^N}\\
&=\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}{1-\left(1-p_w\right)^N}
\end{align}
\begin{align}
P_2\left(\tau_2\right)&=\int_0^\infty \d w P\left(W_l=w\right)\left[1-\e^{-vp_mW_l\left(\tau_2\right)}\right]\\
&=1-\mathbb{E}\left[\e^{-vp_mW_l\left(\tau_2\right)}\right]
\end{align}
\begin{align}
a_\pm=\frac{2-\delta-y\pm\sqrt{\left(\delta+y\right)^2-4y}}{2\left(1-\delta\right)}
\end{align}
\begin{align}
P_2\left(\tau_2\right)=\frac{\left(a_+-1\right)\left(1-a_-\right)\left(1-\e^{\left[-\left(1-\delta\right)\left(a_+-a_-\right)t\right]}\right)}{a_+-1+\left(1-a_-\right)\e^{\left[-\left(1-\delta\right)\left(a_+-a_-\right)t\right]}}
\end{align}
with $y=vp_m$ in $a_\pm$.
\begin{align}
\mathbb{E}\left[\tau_2\right]=\int_0^\infty\left(1-\frac{P_2\left(t\right)}{p_m}\right)\d t
\end{align}
\section{Discussion}
